Calibrating seed-based alignment heuristics with Sesame

### 2020-02-20 manuscript.tex
Start to edit for revisions

\author[1,2]{Guillaume J. Filion}
\author[1]{Ruggero Cortini}
\author[1]{Eduard Zorita}
\affil[1]{Center for Genomic Regulation (CRG), The Barcelona Institute of
Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
\affil[2]{University Pompeu Fabra (UPF), Barcelona, Spain.}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies creates a need
for faster algorithms. The fate of most reads is to be mapped to a
reference sequence, typically a genome. Modern mappers rely on heuristics
to gain speed at a reasonable cost for accuracy. In the seeding heuristic,
short matches between the reads and the genome are used to narrow the
search to a set of candidate locations. Several seeding variants used in
modern mappers show good empirical performance but they are difficult to
calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that the correct
location of a read is filtered out during seeding, resulting in mapping
errors. We describe the properties of simple exact seeds, skip-seeds and
MEM seeds (Maximal Exact Match seeds).
% Innovation.
The main innovation of this work is to use concepts from analytic
combinatorics to represent reads as abstract sequences, and to specify
their generative function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a workable
solution for the problem of calibrating seeding heuristics for short
reads. We also provide a C implementation of these algorithms in a library
called Sesame.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
% Availability
The Sesame library is open source and available for download at
\url{https://github.com/gui11aume/sesame}.
\end{abstract}

@@

Suppose that we use and imperfect instrument to sequence a DNA fragment
drawn at random from a known genome; how can we identify the location of
the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; diagnosing genetic diseases at birth; finding compatible
graft donors \textit{etc}.

At first sight, whether a read can be mapped to its correct location seems
to depend only on its length and on the error rate of the sequencer.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
that are present at more than one location. For instance, we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

This case is rare though. Repeated sequences are in general not identical
but merely homologous --- meaning that their similarity is unlikely to
occur by chance. So when the DNA fragment has one or more duplicates in
the genome, we can still map it, as long as we can distinguish the
duplicates. This in turn depends on their similarity with the target
fragment and on the error rate of the sequencer.

Since repeats play a central role in the problem at hand, we give the
terms ``targets'' and ``duplicates'' a meaning that will facilitate the
exposition of the theory.

\begin{definition}
The target is the DNA fragment that was sequenced. Duplicates are
sequences of the genome that are homologous to the target.
\end{definition}

Because sequencing errors can convert the sequence of the target into one
of its duplicates, the \emph{true} location of a DNA fragment is not
always the \emph{best} candidate (as measured by the identity between the
sequence and the candidate location). This leads to the following
question: how can we identify the \emph{best} location of the fragment in
the genome?

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. For
simplicity we will assume that the true location is always the best, and
we will often confound these two ideas.

@@

There exist algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run faster, but that may
return an incorrect result~\cite{Waterman1984}.

The most popular heuristic for the best location problem is a filtration
method called ``seeding''. The principle is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm at the candidate locations. This is for
instance the general strategy of the popular alignment algorithm
BLAST~\cite{pmid2231712}.

If seeding is fast, we can quickly zoom in on a small set of candidate
locations and reduce the input size of the (slow) alignment algorithm. The
disadvantage is that the target location is not always detected and
included in the candidate set. When that happens, the read cannot be
mapped to the true location because it has been filtered out at the
seeding step.

The slowness of exact alignment algorithms imposes a trade-off between
speed and accuracy: If the seeding step is set to produce many candidates,
it is likely to retrieve the true location. The downside is that
processing all the candidates would take long. Conversely, if the seeding
step is set to produce few candidates, the process would run faster but
the target is less likely to be in the candidate set.

Importantly, the trade-off depends on the seeding method. This means that
we can develop faster mapping algorithms at no cost for accuracy, as long
as we can find better seeding strategies. Progress on this line of
research has largely benefited from the improvement of computer hardware
and from the development of indexing algorithms. Literature on this topic
is already abundant. Some examples of seeding algorithms are given in
references~\cite{sun2005designing, pmid11934743, xu2006optimizing,
kucherov2005multiseed, brejova2003vector, pmid18684737, pmid15359419}.
References~\cite{pmid16533404} and~\cite{pmid20460430} present high-level
comparisons of different designs, and reference~\cite{navarro2001guided}
gives a global overview of filtration methods in pattern matching.

@@

Filtering methods are considered to fail whenever the target is not in the
candidate set. For seeding heuristics, we need to distinguish two
different kinds of failure. In the first kind, the candidate set contains 
a duplicate of the target but does not contain the target itself; in the
second kind, the candidate set contains neither.

The distinction is important because duplicates of the target are similar
to the read (due to their similarity to the target), so a failure of the
first kind is difficult to distinguish from a success. A failure of the
second kind is easier to recognize because in this case the best hit is
not similar to the read, as we will highlight in
section~\ref{sec:random_seeds}.

Before going further, we introduce some nomenclature that will be useful
to simplify the exposition of the theory.

\begin{definition}
The seeding process is said to be
\begin{enumerate}[i)]
\item ``on target'' if the candidate set contains the target,
\item ``off target'' if the candidate set contains a duplicate but
not the target, 
\item ``null'' if the candidate set contains neither.
\end{enumerate}
\end{definition}

If a seed-based mapping algorithm returns an incorrect location, then it
must be the case that the best location was never tested in the
post-filtration phase (otherwise it would have been found by the exact
alignment algorithm).

This observation is fundamental: in order to estimate the error rate of a
mapping algorithm, we must know how often the associated seeding heuristic
is off target. For lack of theoretical framework, this information is
missing, meaning that current mappers are not properly calibrated.

Our focus here is to develop a theory to estimate the probability that
seeding is off target, when using the most popular seeding heuristics for
short reads. Previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to estimate the frequency of mapping errors. Other
authors investigated the reliability of mapping
algorithms~\cite{pmid23872968}, but they focused on the probability that
random sequences have a single hit, recognizing that solving the problem
requires taking into consideration the repeat structure of the genome.
Here we address this problem in general terms and we provide a practical
way to estimate the mismapping rate in complex genomes.

The rest of this document is not organized by logical units but by
increasing level of difficulty. The order of the examples is meant to
ease the acquisition of the main concepts, meaning that we sometimes go
back and forth between different types of seeding strategies. We first
give some background on the different types of seeding methods considered
here; we then introduce the computational strategy based on transfer
matrices; finally we delve into each specific seeding method.

@@

The term ``seed'' changes meaning with different authors. It can designate
a part of the read, a part of the genome, a particular sequence motif or a
structured pattern of matches. Also, the term does not always imply an
exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
uses ``spaced seeds'' that tolerate mismatches. To avoid potential
confusion, we will adopt the following convention:

@@

When a seed matches a particular location of the genome, we say that it is
a ``seed for that location''. For instance, we often use expressions such
as ``a seed for the target'' or ``a seed for a duplicate''.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately we can
use practical methods to solve it, even when the reference genome is very
large.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}. Hash
tables are typically used to index $k$-mers, which makes them useful to
search seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
structures have no set size so they can be used to search seeds of
different lengths. For instance, the FM-index is a compact data structure
based on the suffix array~\cite{manber1993suffix} and on the
Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
with a much smaller memory footprint~\cite{ferragina2000opportunistic,
ferragina2005indexing}.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds, it is even the only
practical option~\cite{pmid24336412, pmid25399029, pmid23349213,
pmid19389736}. However, the detail is of little interest here: we simply
assume that seeds are known at all times without ambiguity because this
problem has several practical solutions.

@@

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
overlapping matches within the same regions.

Fig.~\ref{fig:exact_seed_example} shows a read produced by an imperfect
sequencer. The sequenced DNA fragment comes from a region of the genome
that has three duplicates and the read has three miscalled nucleotides
(indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above the actual molecule
(Target), and above homologous sequences also present in the genome
(Duplicates). Sequencing errors are indicated by a star (each marked
\texttt{T} was actually an \texttt{A} that was misread). The mismatches
between the genomic sequences and the read are indicated in bold. The
exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
above the read. The matching regions in the genomic sequences are
shadowed in grey. Several overlapping seeds accumulate at the center of
the read, which is typical for this seeding method.}
\label{fig:exact_seed_example}
\end{figure}

Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to an off-target seed.
However, sequencing errors that are mismatches for \emph{all} the
sequences cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of seeds.

The middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive compared to spaced seeds~\cite{pmid11934743} so it
is used only in few specific applications. Nevertheless, it will be
extremely useful for the development of the present theory.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
missing seeds are represented by dotted lines.}
\label{fig:skip_seed_example}
\end{figure}

Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
replaced by skip-1 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
seeds. This amounts to removing every second seed. The consequence is that
there are fewer overlapping matches at the center of the read, but the
only seed for the second duplicate is lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but the same might happen to the target.

@@

A \emph{MEM seed} is simply a MEM of size size $\gamma$ or greater.
Fig.~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
read of Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the
center of the read has disappeared because consecutive matches are fused
into a few MEM seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
size $\gamma=7$. The clutter at the center of the read has disappeared and
there is at least one seed for each sequence.}
\label{fig:MEM_example}
\end{figure}

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location using MEM seeds (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
at the bottom effectively hide the target, so it cannot be discovered.
This can occur even when the true location is the best candidate and when
there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
the target and the first duplicate (provided $\gamma \leq 12$).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{short_vs_long_example.pdf}
\caption{\textbf{Issues with MEM seeds: too long reads.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. There would be a good seed (shared) if the
read were two nucleotides shorter. The true location is hidden by
the last two nucleotides.}
\label{fig:short_vs_long}
\end{figure}

@@

Since the decomposition of the read in segments is unique, we can view a
read as a sequence of segments with a tail, instead of a sequence of
nucleotides. Fig.~\ref{fig:simple} shows an example of decomposition in
segments. On-target seeds cannot contain sequencing errors, therefore they
must be completely embedded inside a segment. So the sizes of the segments
indicate whether the read contains an on-target seed or not.

@@

As per definition~\ref{def:seg}, segments in this alphabet are sequences
of 0 or more symbols $\square$ followed by any of the symbols
$\Downarrow_i$ or by the symbol $|$. Given that this decomposition is
unique, we can again view a read as a sequence of segments with a tail.
The example of Fig.~\ref{fig:simple} is shown again in
Fig.~\ref{fig:skip}, where segments in the mismatch alphabet have been
replaced by segments in the skip-3 alphabet.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Fig.~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the symbol $\square$
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$i$ in $\Downarrow_i$ indicates the number of nucleotides till the next
non-skipped position. For $i=0$ the next nucleotide is not skipped.
Other features are as in Fig.~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

Fig.~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
and $-$ sequences are represented below the read, with matches represented
as open squares and mismatches as closed squares. It is visible from this
example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
whenever the mismatches hit different sequences . The symbol $\Downarrow$
occurs only when a nucleotide is a double mismatch.

@@

Fig.~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is clearly visible from expression (\ref{eq:Poff}). This
can also be understood intuitively because the probability of null seeding
decreases when there are more potential candidates, and since the
probability of on-target seeding does not change, the probability of
off-target seeding must increase.

@@

Fig.~\ref{fig:skip_dual} shows the read from Fig.~\ref{fig:dual}
represented in the skip-3 dual encoding. It is important to note several
differences with Fig.~\ref{fig:dual}. The first is that the symbols
$\Downarrow_i$ are not always associated with double mismatches. For
instance, the symbol $\Downarrow_0$ on the right side of the read
corresponds to a mismatch for the $+$ sequence only. This happens whenever
the $+$ and the $-$ sequences are mismatched in the same interval between
non-skipped positions (the mismatches do not need to be on the same
nucleotide).

>>>>

\author[1,2,3]{Guillaume J. Filion}
\author[1]{Ruggero Cortini}
\author[1]{Eduard Zorita}
\affil[1]{Center for Genomic Regulation (CRG), The Barcelona Institute of
Science and Technology, Dr. Aiguader 88, Barcelona 08003, Spain.}
\affil[2]{University Pompeu Fabra (UPF), Barcelona, Spain.}
\affil[3]{Present address: Dept. Biological Sciences, University of
Toronto Scarborough}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies creates a need
for faster algorithms. The fate of most reads is to be mapped to a
reference sequence, typically a genome. Modern mappers rely on heuristics
to gain speed at a reasonable cost for accuracy. In the seeding heuristic,
short matches between the reads and the genome are used to narrow the
search to a set of candidate locations. Several seeding variants used in
modern mappers show good empirical performance but they are difficult to
calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that the correct
location of a read is filtered out during seeding, resulting in mapping
errors. We describe the properties of simple exact seeds, skip-seeds and
MEM seeds (Maximal Exact Match seeds).
% Innovation.
The main innovation of this work is to use concepts from analytic
combinatorics to represent reads as abstract sequences, and to specify
their generative function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which together give a workable solution for
the problem of calibrating seeding heuristics for short reads. We also
provide a C implementation of these algorithms in a library called Sesame.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
% Availability
The Sesame library is open source and available for download at
\url{https://github.com/gui11aume/sesame}.
\end{abstract}

@@

Say we use an imperfect instrument to sequence a small fragment of DNA. If
we know its genome of origin, how can we find the location of the fragment
in this genome?

The advent of high-throughput sequencing has put this question at the
center of countless applications in genetics, such as discovering
disease-causing mutations, selecting breeds of interest in agriculture or
tracing human migrations.

We will refer to this as the \emph{true location problem}. The answer to
the problem depends on the length of the read, on the error rate of the
sequencer and on one key feature of the genome: its repeat structure.
Most genomes contain repetitive sequences, \textit{i.e.}, relatively large
subsequences that are present at multiple locations. If the sequenced DNA
fragment comes from a repetitive sequence, it may be impossible to map the
read with certainty.

In general, repetitive sequences are not identical but merely
homologous---meaning that their similarity is very unlikely to occur by
chance. So if the sequenced DNA fragment originates from a repetitive
sequence, we can map the read only if we can rule out all the candidates
except one. This, in turn, depends on the similarity between the
duplicates and on the error rate of the sequencer.

Since repetitive sequences play a central role in the problem, we give the
terms ``targets'' and ``duplicates'' a meaning that will facilitate the
exposition of the theory.

\begin{definition}
The target is the DNA fragment that was actually sequenced. Duplicates are
sequences of the genome that share homology with the target (in genetics
they are often referred to as paralogs).
\end{definition}

The difficulty of the true location problem is due to sequencing errors.
Occasionally, the sequence of a DNA fragment can be closer to one of the
duplicates than to the target. So the \emph{true} location of a DNA
fragment is not always the \emph{best} (as measured by the identity
between the sequence and the candidate location). This naturally leads to
ask how we can identify the \emph{best} location of the fragment in
the genome.

We will refer to this as the \emph{best location problem}. It amounts to
finding the optimal alignment between two sequences, and for this reason
has received substantial attention in bioinformatics. For the purpose of
developing a theory rooted in statistics, our main concern is to address
the true location problem, but for simplicity, we will assume that the
true location is also the best. When applicable, we will clarify the
implications of this hypothesis in the relevant sections, and we will show
in section~\ref{sec:truevsbest} that this simplifying assumption is harmless for
the present theory.

@@

Exact alignment algorithms were designed to solve the best location
problem~\cite{pmid7265238, pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.}, algorithms that run faster but may
return an incorrect result~\cite{Waterman1984}.

The most popular heuristic for mapping DNA sequences is a filtration
method called seeding. The principle is to first identify seeds,
defined as short regions of high similarity between the read and the
genome, and then use an exact alignment algorithm at the candidate
locations to identify the optimum. This strategy was first proposed in
FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to to search for
homology in sequence databases.

There are efficient methods to extract seeds, so it is possible to quickly
hone in on a small set of candidates and to reduce the search space of the
alignment algorithm. The disadvantage is that the target may not be in the
candidate set, in which case the read cannot be mapped correctly.

As a consequence, seeding methods induce a trade-off between speed and
accuracy: If the filtration is set to produce a large candidate set, the
target is likely to be discovered, with the downside that checking all the
candidates with an exact alignment algorithm will take long. Conversely,
if the filtration is set to produce a small candidate set, the process
will run faster but the target is more likely to be missed.

Importantly, the trade-off depends on the seeding method. This means that
mapping algorithms can run faster at no cost for accuracy if we can find
better seeding strategies. Progress on this line of research has largely
benefited from the improvement of computer hardware and from the
development of optimized data structures. There already exists a large
body of literature on the design of seeding algorithms; the interested
reader can find examples of those in references~\cite{sun2005designing,
pmid11934743, xu2006optimizing, kucherov2005multiseed, brejova2003vector,
pmid18684737, pmid15359419}. References~\cite{pmid16533404}
and~\cite{pmid20460430} present high-level comparisons of different
designs, and reference~\cite{navarro2001guided} gives a global overview of
filtration methods in pattern matching.

@@

Filtering heuristics are considered to fail whenever the target is not in
the candidate set, but here we must be more specific and distinguish two
kinds of failure: In the first kind, the candidate set contains a
duplicate of the target but does not contain the target itself; in the
second kind, the candidate set contains neither the target nor any
duplicate.

The distinction is important because the duplicates of the target are
similar to the read (due to their similarity to the target), so a failure
of the first kind often looks like a success. In contrast, a failure of
the second kind is easier to flag because in this case the candidates are
not similar to the read. We will simply assume that seeding failures of
the second kind are always detected (as explained in
section~\ref{sec:random_seeds}), so that we can focus on the more
difficult case of seeding failures of the first kind.

Before going further, we introduce three terms that will simplify the
exposition.

\begin{definition}
The candidate set obtained at the seeding step is said to be
\begin{enumerate}[i)]
\item ``on target'' if it contains the target,
\item ``off target'' if it contains a duplicate but not the target, 
\item ``null'' if it contains neither.
\end{enumerate}
\end{definition}

With our assumptions, a read is mapped to the wrong location if and only
if the candidate set is off target. Indeed, if the candidate set is null,
the read is not mapped, and if the candidate set is on target, the correct
location is discovered at the alignment step. The equivalence is granted
by the assumption that the true location is also the best, reducing the
mapping problem to a seeding problem. That being said, the true location
is not always the best in practical applications. We will show in
section~\ref{sec:truevsbest} that off-target seeding is responsible for
most of the mapping errors even without the assumption above, but for now
we maintain the strict equivalence between mapping errors and off-target
seeding.

Focusing on popular heuristics to map short reads, our aim is to develop a
method to estimate the probability that the candidate set is off target.
Previous work pioneered a method to compute seeding probabilities but it
did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to estimate the frequency of mapping errors. Other
authors investigated the reliability of mapping
algorithms~\cite{pmid23872968}, but they focused on the probabilities of
random hits, recognizing that addressing the problem of incorrect mapping
requires taking into consideration the repeat structure of the genome.

The rest of this document is organized so as to present the concepts by
increasing level of complexity, meaning that we sometimes go back and
forth between different types of seeding strategies. We first provide some
background knowledge on different types of seeding methods, we then
introduce the computational strategy based on transfer matrices, and
finally we delve into each specific seeding method.

@@

The term ``seed'' has different meanings in computational biology. It can
designate a part of the read, a part of the genome, a particular sequence
motif, or a structured pattern of matches. Also, a seed does not always
refer to an exact match. For instance, the algorithm
PatternHunter~\cite{pmid11934743} uses ``spaced seeds'' that tolerate
mismatches. To avoid any confusion, we will adopt the convention below.

@@

When a seed matches a given location of the genome, we say that it is a
seed for that location. This is particularly useful in expressions such as
``seed for the target'' or ``seed for a duplicate''.

This definition presents a computational challenge: to know if a given
subsequence of a read is a seed we need to know if it exists somewhere in
the genome. This is a non trivial problem in itself, but fortunately we
can use practical methods to solve it, even when the reference genome is
very large.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{ferragina2000opportunistic, ferragina2005indexing}. Hash
tables are typically used to index $k$-mers, which makes them useful to
search for seeds of fixed size $k$ (see~\cite{pmid30346548} for a recent
benchmark of $k$-mer hashing algorithms). In contrast, some text-indexing
structures have no set size so they can be used to search for seeds of
different lengths. For instance, the FM-index is a compact data structure
based on the suffix array~\cite{manber1993suffix} and on the
Burrows-Wheeler transform~\cite{burrows1994block}, emulating a suffix trie
with a much smaller memory footprint~\cite{ferragina2000opportunistic,
ferragina2005indexing}.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds defined below, it is
even the only practical option~\cite{pmid24336412, pmid25399029,
pmid23349213, pmid19389736}. Overall, the detail is of little interest for
the theory. We simply assume that seeds are known at all times without
ambiguity because this problem has several practical solutions.

@@

Exact seeds are seeds of fixed size $\gamma$. In other words, when using
exact seeds, the candidate set consists of all the genomic locations for
which there is a perfect match of size $\gamma$ in the read. This seeding
heuristic was used in the first version of BLASTN~\cite{pmid2231712}, but
it has become unpopular for producing many redundant hits within the same
regions of the read and of the genome.

Figure~\ref{fig:exact_seed_example} shows the exact seeds from an example
a read with three miscalled nucleotides. The sequenced DNA fragment has
three duplicates so the seeds can match four possible locations.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Exact seeding.}
A sequenced DNA fragment (Read) is shown above the actual molecule
(Target), and above three paralogs of the target (Duplicates). Sequencing
errors are indicated by stars (all three are \texttt{A} misread as
\texttt{T}). The mismatches between the genomic sequences and the read are
indicated in bold. The exact seeds of size $\gamma = 7$ are indicated as
horizontal grey lines above the read. Matching regions in the genomic
sequences are shadowed in grey. Several overlapping seeds accumulate at
the center of the read, which is typical of exact seeds.}
\label{fig:exact_seed_example}
\end{figure}

Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match a duplicate. For instance, the first sequencing error
matches all the duplicates and belongs to an off-target seed for the first
duplicate. However, sequencing errors that are mismatches for \emph{all}
the sequences cannot belong to a seed. This is the case of the second
sequencing error in the example, creating a local deficit of seeds.

Note the clutter in the middle of the read, where consecutive seeds match
consecutive sequences at the same location. This is typical for exact
seeds and is considered a nuisance for the implementation. Indeed, it is a
waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive compared to spaced seeds~\cite{pmid11934743} so it
is used only in a few specific applications. Nevertheless, it will be
useful for the development of the present theory.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Figure~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
highlight the difference with Figure~\ref{fig:exact_seed_example}, the
missing seeds are represented by dotted lines.}
\label{fig:skip_seed_example}
\end{figure}

Figure~\ref{fig:skip_seed_example} shows what happens when exact seeds are
replaced by skip-1 seeds on the read of Figure~\ref{fig:exact_seed_example}.
Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
seeds. This amounts to removing every second seed. The consequence is that
there are fewer overlapping matches at the center of the read, but the
only seed for the second duplicate is lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but the same might happen to the target.

@@

A \emph{MEM seed} is simply a MEM of size size $\gamma$ or greater.
Figure~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
read of Figure~\ref{fig:exact_seed_example}. Observe that the clutter at the
center of the read has disappeared because consecutive matches are fused
into a few MEM seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=1]{MEM_example.pdf}
\caption{\textbf{Example of MEM seeds.}
The sequences and the annotations are the same as in
Figure~\ref{fig:exact_seed_example}, but here we use MEM seeds of minimum
size $\gamma=7$. The clutter at the center of the read has disappeared and
there is at least one seed for each sequence.}
\label{fig:MEM_example}
\end{figure}

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Figure~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location using MEM seeds (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Figure~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
at the bottom effectively hide the target, so it cannot be discovered.
This can occur even when the true location is the best candidate and when
there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Figure~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
the target and the first duplicate (provided $\gamma \leq 12$).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{short_vs_long_example.pdf}
\caption{\textbf{Issues with MEM seeds: too long reads.}
The read, the MEM seeds and the sequences are represented as in
Figure~\ref{fig:MEM_example}. There would be a good seed (shared) if the
read were two nucleotides shorter. The true location is hidden by
the last two nucleotides.}
\label{fig:short_vs_long}
\end{figure}

@@

Since the decomposition of the read in segments is unique, we can view a
read as a sequence of segments with a tail, instead of a sequence of
nucleotides. Figure~\ref{fig:simple} shows an example of decomposition in
segments. On-target seeds cannot contain sequencing errors, therefore they
must be completely embedded inside a segment. So the sizes of the segments
indicate whether the read contains an on-target seed or not.

@@

As per definition~\ref{def:seg}, segments in this alphabet are sequences
of 0 or more symbols $\square$ followed by any of the symbols
$\Downarrow_i$ or by the symbol $|$. Given that this decomposition is
unique, we can again view a read as a sequence of segments with a tail.
The example of Figure~\ref{fig:simple} is shown again in
Figure~\ref{fig:skip}, where segments in the mismatch alphabet have been
replaced by segments in the skip-3 alphabet.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_i (i = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the symbol $\square$
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$i$ in $\Downarrow_i$ indicates the number of nucleotides till the next
non-skipped position. For $i=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

Figure~\ref{fig:dual} shows an example of read in the dual encoding. The $+$
and $-$ sequences are represented below the read, with matches represented
as open squares and mismatches as closed squares. It is visible from this
example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
whenever the mismatches hit different sequences . The symbol $\Downarrow$
occurs only when a nucleotide is a double mismatch.

@@

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is clearly visible from expression (\ref{eq:Poff}). This
can also be understood intuitively because the probability of null seeding
decreases when there are more potential candidates, and since the
probability of on-target seeding does not change, the probability of
off-target seeding must increase.

@@

Figure~\ref{fig:skip_dual} shows the read from Figure~\ref{fig:dual}
represented in the skip-3 dual encoding. It is important to note several
differences with Figure~\ref{fig:dual}. The first is that the symbols
$\Downarrow_i$ are not always associated with double mismatches. For
instance, the symbol $\Downarrow_0$ on the right side of the read
corresponds to a mismatch for the $+$ sequence only. This happens whenever
the $+$ and the $-$ sequences are mismatched in the same interval between
non-skipped positions (the mismatches do not need to be on the same
nucleotide).
